\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2019-02-07 document/XSGRB_sample_rev1.tex
Accepted version

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. In total, we obtained spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneously selected sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We constrain the fraction
	of dark bursts to be < 28 per cent and we confirm previous results that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we provide a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening universe.}

@@

We remark, however, that while the selection of the afterglows has been carried
out in a homogeneous way, the resulting quality of the spectra is not uniform.
Individual bursts vary significantly in brightness, due to intrinsic
differences in luminosity, dust extinction, time of observation, etc. The covered
rest-frame wavelength range is also different for each case. Given that our
observations are targets-of-opportunity, this leads to a varying S/N, and in
particular the sensitivity limit to detecting absorption features varies greatly
from burst to burst and as a function of wavelength.

@@

The first GRB followed up was GRB~090313 \citep{DeUgartePostigo2010}, observed
on the 15th of March, 2009, during the commissioning of X-shooter on the second
Unit Telescope (UT2) of the VLT . The bursts observed during the commissioning
or the science verification process (GRB~090313, GRB~090530, GRB~090809,
GRB~090926A) are not a part of the sample we use to investigate the statistical
properties of GRB afterglows, due to the different criteria for their selection. The
first burst observed after science verification and the mounting of X-shooter on UT2,
was GRB~091018, which is the first burst entering our statistically homogeneously selected
sample. For all bursts that fulfill our sample selection criteria, described in
Sect. \ref{samplecrit}, spectroscopic follow-up have been attempted with
X-shooter. Various conditions can affect our ability to follow up a given burst,
and a discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

The resulting offsets are generally small, on the order of $\sim 5$~km~s$^{-1}$,
and thus do not influence significantly the redshift measurements, whose
uncertainties are primarily driven by the S/N of the spectra, the line-shapes,
and the choice of ion species. Such a correction, however, can improve the
accuracy of line fits, especially when multiple components are tied together,
across a wide wavelength range, and allows a better comparison with
high-resolution data. The scale of the correction additionally serves to
solidify the accuracy of the wavelength solution.

@@

To correct for the slit-loss, we develop an algorithm that can, based on the
atmospheric conditions at time of observation and the wavelength dependence of
seeing, estimate the losses at each wavelength bin. We start with the DIMM
seeing \citep{Sarazin1990}, which measures the seeing FWHM at zenith. We correct
the zenith seeing for the airmass of the observational sightline, which gives us
the effective seeing, delivered outside the slit at 5000 \AA. We can then use
that the seeing scales as $\sim \lambda^{-1/5}$ to obtain a measure of the
seeing across the entire wavelength coverage of X-shooter. At each wavelength
bin of the spectra, we simulate a synthetic 2D seeing PSF based on a normalised,
2D Moffat profile which we integrate over the used slit width. This can be used to correct for the estimated slit-loss at each wavelength bin.

@@

To compare the photometry with the spectral flux calibration,
we calculate synthetic magnitudes from the spectrum, in the used passbands,
along with the associated error. Using this set we can make a test of the final flux-calibration accuracy, which
we show in Fig.~\ref{fig:flux_acq}. As can be seen, there is a general agreement
between the flux-calibration and the derived acquisition magnitudes, although
with significant scatter. The right panel in Fig.~\ref{fig:flux_acq} shows a
histogram of the differences, which can be summarized with $\Delta \mathrm{mag}
= 0.07 ^{+0.51} _{-0.45} $, where the errors are the 16th and 84th percentiles of
the distribution. The scatter is smaller if we exclude measurements calibrated against the USNO catalog, which is known to have a poor accuracy, also because, given the small field-of-view, we often need to rely of faint reference stars.

@@

The effect of telluric correction varies greatly depending on the wavelength
region. In the red and near-infrared areas, it allows to identify features in a
credible manner. which would be otherwise hidden among or blended with
atmospheric features. The telluric correction allows us to reliably recover the
continuum level even in regions of moderately low atmospheric transmission (see,
e.g., the A- and B-bands in Fig.~\ref{fig:spectrum}). Nonetheless, we recommend
caution when fitting lines in regions significantly affected by telluric
absorption.

@@

First, numerous works, including this one, have found that GRB absorption
systems have much higher column densities than randomly selected sight lines
\citep{2001ApJ...549L.209G, Savaglio2003,  Prochaska2007, Fynbo2009}. This fact
itself shows that the bulk of GRB absorption lines are not due to random
intervening systems, but are peculiar to the GRB host. A second line of evidence
comes from the systematic detection of nebular emission lines at the same
redshift as the absorption system. \citet{Kruhler2015} report a 94\% detection
rate of emission lines at $z < 1$ (where sensitivity is not an issue, even in
the presence of a bright afterglow), always in agreement with absorption.
Besides, it is rare to detect emission from QSO intervening systems
(see e.g. \citealt{2010MNRAS.408.2128F} for a discussion), which again indicates
that GRB absorption systems are native to their hosts. An even more direct
determination comes from the detection, in several cases, of fine-structure and
metastable lines (for example from Fe and Ni), which are exclusively detected in
GRB environments (and have been directly connected to the UV field generated by
the GRB itself; e.g. \citealt{2006ApJ...648...95P, Vreeswijk2007, DElia2009}).
Finally, at $z \gtrsim 2$ (where Ly$\alpha$ enters the observed range) the onset
of the Ly$\alpha$ forest also provides an \textit{upper} limit on the GRB
redshift, which again is always consistent with the absorption line redshift.
The whole set of these arguments indicates that in only very few cases, if any
at all, is a misidentification of the redshift with the intervening system a
concrete possibility.

@@

\begin{figure}[!t]
	\centering \includegraphics[width=\columnwidth]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citet{Noterdaeme2012b} in blue. Overplotted in
	green is the kernel density estimate of absorbers in GRB sightlines. Values are
	taken from the compilation in \citet{Tanvir2017}, along with the new
	values presented in this sample. We also show in the red the values derived in
	this work. The marginal distributions for the three samples are also shown
	along the left side and on the top, where the different environments probed are
	clearly visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}. To correct for geometric differences between GRB DLAs and QSO DLAs, we have multiplied the inferred GRB DLA $N_{\mathrm{HI}}$ by 2.} \label{fig:NH_dist}
\end{figure}

@@

\begin{acknowledgements}
Much of the work done here would not have been done without Neil Gehrels, who
has now passed away. We owe him a large debt for his invaluable work with the
\swift~satellite. We also want to remember Javier Gorosabel, who was taken from
us too early. 
%
JPUF, BMJ and DX acknowledge support from the ERC-StG grant EGGS-278202.
The Dark Cosmology Centre was funded by the Danish National Research
Foundation. 
%
This work was supported by a VILLUM FONDEN Investigator grant to JH
(project number 16599). 
%
TK acknowledges support by the European Commission
under the Marie Curie Intra-European Fellowship Programme in FP7.  
%
LK and JJ acknowledges
support from NOVA and NWO-FAPESP grant for advanced instrumentation in
astronomy. 
%
KEH and PJ acknowledge support by a Project Grant (162948--051) from
The Icelandic Research Fund. 
%
AG acknowledges the financial support from the
Slovenian Research Agency (research core funding No. P1-0031 and project grant
No. J1-8136). 
%
CT acknowledges support from a Spanish National Research Grant of Excellence
under project AYA 2014-58381-P and funding associated to a Ramón y Cajál
fellowship under grant number RyC-2012-09984.
%
AdUP acknowledges support from a Ramón y Cajal fellowship, a BBVA Foundation
Grant for Researchers and Cultural Creators, and the Spanish Ministry of Economy
and Competitiveness through project AYA2014-58381-P.
%
ZC acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2014-21669.
%
DAK acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2015- 26153.
%
RSR acknowledges AdUP's BBVA Foundation Grant for Researchers and Cultural
Creators and support from ASI (Italian Space Agency) through the Contract n.
2015-046-R.0 and from European Union Horizon 2020 Programme under the AHEAD
project (grant agreement n. 654215).
%
GL is supported by a research grant (19054) from VILLUM FONDEN.
%
SDV acknowledges the support of the French National Research Agency (ANR) under
contract ANR-16-CE31-0003 BEaPro
%
DM acknowledges support from the Instrument Center for Danish Astrophysics
(IDA).
%
This work made use of data supplied by the UK {\it Swift} Science Data Centre at
the University of Leicester. Finally, it is with pleasure that we acknowledge
expert support from the ESO staff at the Paranal and La Silla observatories in
obtaining these target of opportunity data.
%
This research made use of Astropy, a community-developed core Python package for
Astronomy \citep{TheAstropyCollaboration2013}. The analysis and plotting was
achieved using the Python-based packages Matplotlib \citep{Hunter2007}, Numpy,
and Scipy \citep{scipy, VanderWalt2011}, along with other community-developed
packages.
%

@@

\bibliographystyle{aa}
\bibliography{XSGRB_sample,GCN}

@@

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. The black dashed line shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines. The pressence of \lya, as noted in Tab. \ref{tab:HI}, is indicated with a black arrow.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=6, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI6}
\end{figure*}
\clearpage

@@

These data have not been published elsewhere. Observations for GRB~161007A
started 323 hr after the burst trigger and were centred on the potential host
galaxy (see GCN 20014 \citep{GCN20014} and GCN 20020 \citep{GCN20020}). $4
\times 600$ s of observations were obtained revealing a faint continuum rising
abruptly above the noise at $\sim 685$ nm and extending up to 2100 nm. A very
low significance continuum is detected at shorter wavelengths, down to $\sim
600$ nm. No significant emission features could be identified. Two possibilities
can explain the observed continuum break, but neither are fully satisfactory.
The first option is the Lyman alpha cutoff at $z \sim 4.6$; at this redshift, the
usually detected strong nebular lines would be shifted out of the covered
wavelength range. The host would be however exceptionally bright, with an
absolute magnitude at $\approx 150$~nm{} $M = -23.8$~AB, that is about 2.8~mag
brighter than $M^*$ at that redshift (e.g. \citealt{Bouwens2015}). The inferred
X-ray hydrogen-equivalent column density would also be a record-high $4 \times
10^{23}$~cm$^{-2}$ (see e.g. \citealt{Campana2012}). Alternatively, the
continuum discontinuity could be the 400~nm{} break at $z = 0.71$; such
feature is however prominent only in early-type galaxies, which are unlikely
hosts of long-duration GRBs (but see \citealt{Rossi2014}). Given the lack of
optical variability, it cannot be excluded that the object within the XRT error
circle is a chance association, as the probability of a random alignment is
small but non negligible (of the order of 1\%). In conclusion, we cannot provide
a secure redshift for GRB 161007A.

>>>>

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. In total, we have obtained spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneously selected sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We constrain the fraction
	of dark bursts to be < 28 per cent and confirm previous results that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we provide a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening Universe.}

@@

We remark, however, that while the selection of the afterglows has been carried
out in a homogeneous way, the resulting quality of the spectra is not uniform.
Individual bursts vary significantly in brightness, due to intrinsic differences
in luminosity, dust extinction, time of observation, etc. The covered rest-frame
wavelength range is also different for each case due to the different redshifts
of each GRB. Given that our observations are targets-of-opportunity, this leads
to a varying S/N, and in particular the sensitivity limit to detect absorption
features varies greatly from burst to burst and as a function of wavelength.

@@

The first GRB followed up was GRB~090313 \citep{DeUgartePostigo2010}, observed
on the 15th of March, 2009, during the commissioning of X-shooter on the second
Unit Telescope (UT2) of the VLT . The bursts observed during the commissioning
or the science verification process (GRB~090313, GRB~090530, GRB~090809,
GRB~090926A) are not a part of the sample we use to investigate the statistical
properties of GRB afterglows, due to the different criteria for their selection.
The first burst observed after science verification and the mounting of
X-shooter on UT2, was GRB~091018, which is the first burst entering our
homogeneously selected statistical sample. For all bursts that fulfill our
sample selection criteria, described in Sect. \ref{samplecrit}, spectroscopic
follow-up have been attempted with X-shooter. Various conditions can affect our
ability to follow up a given burst, and a discussion of these conditions and
their consequences for the sample is included in Sect. \ref{badbursts}.

@@

The resulting offsets are generally small, on the order of $\sim 5$~km~s$^{-1}$,
and thus do not influence significantly the redshift measurements, whose
uncertainties are primarily driven by the S/N of the spectra, the line-shapes,
and the choice of ion species. Such a correction, however, can improve the
accuracy of line fits, especially when multiple components are tied together,
across a wide wavelength range, and allows a better comparison with
high-resolution data. The modest size of the correction additionally serves to
confirm the accuracy of the wavelength solution.

@@

To correct for the slit-loss, we developed an algorithm that can, based on the
atmospheric conditions at the time of observation and the wavelength dependence
of seeing, estimate the losses at each wavelength bin. We start with the DIMM
seeing \citep{Sarazin1990}, which measures the seeing FWHM at zenith. We correct
the zenith seeing for the median airmass of the observational sightline, which
gives us the effective seeing, delivered at 5000 \AA. We then use the fact that the
seeing scales as $\sim \lambda^{-1/5}$ to obtain a measure of the seeing across
the entire wavelength coverage of X-shooter. At each wavelength bin of the
spectra, we simulate a synthetic 2D seeing PSF based on a normalised, 2D Moffat
profile which we integrate over the used slit width. This can be used to correct
for the estimated slit-loss at each wavelength bin.

@@

To compare the photometry with the spectral flux calibration,
we calculate synthetic magnitudes from the spectrum, in the used passbands,
along with the associated error. Using this set we can make a test of the final flux-calibration accuracy, which
we show in Fig.~\ref{fig:flux_acq}. As can be seen, there is a general agreement
between the flux-calibration and the derived acquisition magnitudes, although
with significant scatter. The right panel in Fig.~\ref{fig:flux_acq} shows a
histogram of the differences, which can be summarized with $\Delta \mathrm{mag}
= 0.07 ^{+0.51} _{-0.45} $, where the errors are the 16th and 84th percentiles of
the distribution. The scatter is smaller if we exclude measurements calibrated against the USNO catalog, which is known to have a poor accuracy, also because, given the small field-of-view, we often need to rely on faint reference stars.

@@

The effect of telluric correction varies greatly depending on the wavelength
region. In the red and near-infrared ranges, telluric correction allows to
identify features in a credible manner, which would be otherwise hidden among or
blended with atmospheric features. The telluric correction also allows us to reliably
recover the continuum level even in regions of moderately low atmospheric
transmission (see, e.g., the A- and B-bands in Fig.~\ref{fig:spectrum}).
Nonetheless, we recommend caution when fitting lines in regions significantly
affected by telluric absorption.

@@

First, numerous works, including this one, have found that GRB absorption
systems have much higher column densities than randomly selected sight lines
\citep{2001ApJ...549L.209G, Savaglio2003,  Prochaska2007, Fynbo2009}. This fact
itself shows that the bulk of GRB absorption lines are not due to random
intervening systems, but are peculiar to the GRB host. A second line of evidence
comes from the systematic detection of nebular emission lines at the same
redshift as the absorption system. \citet{Kruhler2015} report a 94\% detection
rate of emission lines at $z < 1$ (where sensitivity is not an issue, even in
the presence of a bright afterglow), always in agreement with absorption.
Besides, it is rare to detect emission from QSO intervening systems (see e.g.
\citealt{2010MNRAS.408.2128F} for a discussion), which again indicates that GRB
absorption systems are native to their hosts. An even more direct determination
comes from the detection, in several cases, of fine-structure and metastable
lines (for example from Fe and Ni), which are usually detected in GRB
environments (and have been directly connected to the UV field generated by the
GRB itself; e.g. \citealt{2006ApJ...648...95P, Vreeswijk2007, DElia2009}).
Finally, at $z \gtrsim 2$ (where Ly$\alpha$ enters the observed range) the onset
of the Ly$\alpha$ forest also provides confirming evidence of the GRB
redshift, which again is always consistent with the absorption line redshift.
The whole set of these arguments indicates that in only very few cases, if any
at all, is a misidentification of the redshift with the intervening system a
concrete possibility.

@@

\begin{figure}[!t]
	\centering \includegraphics[width=\columnwidth]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citet{Noterdaeme2012b} in blue. Overplotted in
	green is the kernel density estimate of absorbers in GRB sightlines. Values are
	taken from the compilation in \citet{Tanvir2017}, along with the new
	values presented in this sample. We also show in the red the values derived in
	this work. The marginal distributions for the three samples are also shown
	along the left side and on the top, where the different environments probed are
	clearly visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}. To correct for geometric differences between GRB DLAs and QSO DLAs, we have multiplied the inferred GRB DLA $N_{\mathrm{HI}}$ by a factor of two} \label{fig:NH_dist}
\end{figure}

@@

\begin{acknowledgements}
Much of the work done here would not have been done without Neil Gehrels, who
has now passed away. We owe him a large debt for his invaluable work with the
\swift~satellite. We also want to remember Javier Gorosabel, who was taken from
us too early. 
%
We thank Andrea Rossi (INAF/OAS) for the help with providing a photometric
constraint on the redshift of GRB~161007A.
%
JPUF, BMJ and DX acknowledge support from the ERC-StG grant EGGS-278202.
The Dark Cosmology Centre was funded by the Danish National Research
Foundation. 
%
This work was supported by a VILLUM FONDEN Investigator grant to JH
(project number 16599). 
%
TK acknowledges support by the European Commission
under the Marie Curie Intra-European Fellowship Programme in FP7.  
%
LK and JJ acknowledges
support from NOVA and NWO-FAPESP grant for advanced instrumentation in
astronomy. 
%
KEH and PJ acknowledge support by a Project Grant (162948--051) from
The Icelandic Research Fund. 
%
AG acknowledges the financial support from the
Slovenian Research Agency (research core funding No. P1-0031 and project grant
No. J1-8136). 
%
CT acknowledges support from a Spanish National Research Grant of Excellence
under project AYA 2014-58381-P and funding associated to a Ramón y Cajál
fellowship under grant number RyC-2012-09984.
%
AdUP acknowledges support from a Ramón y Cajal fellowship, a BBVA Foundation
Grant for Researchers and Cultural Creators, and the Spanish Ministry of Economy
and Competitiveness through project AYA2014-58381-P.
%
ZC acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2014-21669.
%
DAK acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2015- 26153.
%
RSR acknowledges AdUP's BBVA Foundation Grant for Researchers and Cultural
Creators and support from ASI (Italian Space Agency) through the Contract n.
2015-046-R.0 and from European Union Horizon 2020 Programme under the AHEAD
project (grant agreement n. 654215).
%
GL is supported by a research grant (19054) from VILLUM FONDEN.
%
SDV acknowledges the support of the French National Research Agency (ANR) under
contract ANR-16-CE31-0003 BEaPro
%
DM acknowledges support from the Instrument Center for Danish Astrophysics
(IDA).
%
This work made use of data supplied by the UK {\it Swift} Science Data Centre at
the University of Leicester. Finally, it is with pleasure that we acknowledge
expert support from the ESO staff at the Paranal and La Silla observatories in
obtaining these target of opportunity data.
%
This research made use of Astropy, a community-developed core Python package for
Astronomy \citep{TheAstropyCollaboration2013}. The analysis and plotting was
achieved using the Python-based packages Matplotlib \citep{Hunter2007}, Numpy,
and Scipy \citep{scipy, VanderWalt2011}, along with other community-developed
packages.
%
This publication makes use of data products from the Wide-field Infrared Survey Explorer, which is a joint project of the University of California, Los Angeles, and the Jet Propulsion Laboratory/California Institute of Technology, funded by the National Aeronautics and Space Administration.
%
The Pan-STARRS1 Surveys (PS1) and the PS1 public science archive have been made possible through contributions by the Institute for Astronomy, the University of Hawaii, the Pan-STARRS Project Office, the Max-Planck Society and its participating institutes, the Max Planck Institute for Astronomy, Heidelberg and the Max Planck Institute for Extraterrestrial Physics, Garching, The Johns Hopkins University, Durham University, the University of Edinburgh, the Queen's University Belfast, the Harvard-Smithsonian Center for Astrophysics, the Las Cumbres Observatory Global Telescope Network Incorporated, the National Central University of Taiwan, the Space Telescope Science Institute, the National Aeronautics and Space Administration under Grant No. NNX08AR22G issued through the Planetary Science Division of the NASA Science Mission Directorate, the National Science Foundation Grant No. AST-1238877, the University of Maryland, Eotvos Lorand University (ELTE), the Los Alamos National Laboratory, and the Gordon and Betty Moore Foundation.
%

@@

\newpage
\bibliographystyle{aa}
\bibliography{XSGRB_sample,GCN}

@@

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. The black dashed line shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines. The presence of \lya, as noted in Tab. \ref{tab:HI}, is indicated with a black arrow.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=6, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI6}
\end{figure*}
\clearpage

@@

These data have not been published elsewhere. Observations for GRB 161007A
started 323 hr after the burst trigger and targeted the potential host galaxy
(see GCN 20014 \citep{GCN20014} and GCN 20020 \citep{GCN20020}). $4 \times 600$s
of observations were obtained, revealing a faint continuum rising abruptly above
the noise at $\sim 685$~nm and extending up to 2100 nm. A very low significance
continuum is detected at shorter wavelengths, down to $\sim  600 $ nm. No
significant emission features could be identified. Two possibilities can explain
the observed continuum break, but neither is fully satisfactory. The first
option is the Ly$\alpha$ cutoff at $z \sim 4.6$; at this redshift, the usually detected
strong nebular lines would be shifted out of the covered wavelength range. The
host would be however exceptionally bright, with an absolute magnitude at
$\approx 150$~nm{} $M = -23.8$~AB, that is about 2.8 mag brighter than $M^*$ at
that redshift (e.g. \citealt{Bouwens2015}). In order to test this option, we
secured $B$-band imaging (using the ESO/VLT UT1 equipped with the FORS2 camera)
of the host, yielding a faint but significant detection ($B = 25.48 \pm 0.13$
Vega). As the $B$ band would lie entirely bluewards of the Lyman limit at $z =
4.6$, where negligible flux is expected to be transmitted by the IGM, our
detection rules out this solution. A second solution is that the continuum
discontinuity is the 400 nm break at $z = 0.71$, such feature is however more
prominent in evolved galaxies, which are unlikely hosts of long-duration GRBs
(but see \citealt{Rossi2014}). A fit to the sparse available photometry,
including our FORS2 datum, the GTC $i$-band value \citep{GCN20014}, Pan-STARRS
measurements \citep{Chambers2016}, and WISE \citep{Wright2010} W1 detection
($\mbox{W1}=19.72\pm0.15$), is unable to tightly constrain the redshift ($0.47 <
z < 1.34$). Fixing $z = 0.7$ yields however a significant star formation rate
(25--200 $M_\odot$~yr$^{-1}$), which is in contrast with the lack of detected
emission lines in the spectrum (the expected H$\alpha$ flux would be
$10^{-15}$--$10^{-16}$ erg cm$^{-2}$ s$^{-1}$, easily detectable). As such we
cannot identify a self-consistent solution. Given the lack of optical
variability of the candidate counterpart, it cannot be excluded that the object
within the XRT error circle is a chance association, as the probability of a
random alignment is small but non negligible (of the order of 1\%). In
conclusion, we cannot provide a secure redshift for GRB 161007A.
